A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-31 manuscript.qmd
Built site for gh-pages

Population parameters are all model parameters estimated in a GLMM, in particular the regression coefficients of the fixed effects and the standard deviation of the random effects (and the correlation between random effects in more complicated models).
In the absence of previous studies with the same design or pilot data, strategies to specify population parameters will always require access to some domain knowledge. In our experience, research teams for projects where tailored simulation-based sample size planning is necessary typically consist of *analysts* (who are primarily responsible for study design and statistical analysis), and *domain experts* (who often provide the original research question and are responsible for data collection).
In smaller projects, analysts and domain experts might be the same persons. In these research settings, a broad review of the available literature typically plays a more important role.
Because most domain knowledge can only be expressed in unstandardized measurement units of a specific application, we argue that unstandardized effect sizes are usually preferable over standardized effect sizes for tailored simulation-based sample size planning. The basic idea of all strategies is to quantify or visualize the data-generating process implied by certain values of population parameters in an intuitive way that enables calibration of population parameters based on the available domain knowledge.
Although we use frequentist model estimation in our tutorial, many strategies demonstrated in this chapter are inspired by research on monitoring the plausibility of model assumptions [@gelmanBayesianWorkflow2020; @schadPrincipledBayesianWorkflow2021] and eliciting prior information [@bocktingSimulationBasedPriorKnowledge2023; @mikkolaPriorKnowledgeElicitation2023; @hartmannFlexiblePriorElicitation2020; @stefanPracticalChallengesMethodological2022] in applied Bayesian statistics.

@@

Tailored simulation-based sample size planning always requires access to domain knowledge. However, gathering domain knowledge and the specification of population parameter values is not always straightforward and can be better described as an iterative process, in which the plausibility of simulated data is repeatedly compared with the available domain expertise. As a first step, dedicated domain experts (if available) can be interviewed to *elicit* their domain knowledge about how the data of the planned study is expected to look. Because most domain experts are no experts in statistical modeling and GLMMs, they often struggle without further guidance to communicate their knowledge in a way that is useful when specifying the parameters for data simulation. For this reason, we suggest that after an initial interview, the analyst who is familiar with the structure of the GLMM selects an initial set of insightful descriptive statistics and model-based quantities. Then, they reenter into an iterative discussion with the domain experts where some set of population values are selected, and the plausibility of resulting implied quantities is evaluated. The population parameters are updated until the domain experts are satisfied with the result. During this process, the monitored descriptive statistics and model-based quantities can be updated or extended to capture as much available domain knowledge as possible.
Even if no dedicated domain experts are available, the basic principles stay the same except that initial parameter values are often derived from a more extensive literature review. Descriptive statistics and model-based quantities are still used to evaluate whether the implied assumptions about the data generating process are plausible. Often, this judgment evolves over several iterations as the analyst becomes more familiar with the model and the study design.^[Developing more concrete strategies on how to elicit and visualize domain knowledge is currently an active area of research, primarily in the context of Bayesian modeling [for reviews, see for example @mikkolaPriorKnowledgeElicitation2023 and @stefanPracticalChallengesMethodological2022]. We expect that future advances in this field will also be highly relevant for tailored simulation-based sample size planning with frequentist models.]

>>>>

Population parameters are all model parameters estimated in a GLMM, in particular the regression coefficients of the fixed effects and the standard deviation of the random effects (and the correlation between random effects in more complicated models).
In the absence of previous studies with the same design or pilot data, strategies to specify population parameters will always require access to some domain knowledge. In our experience, research teams for projects where tailored simulation-based sample size planning is necessary typically consist of *analysts* (who are primarily responsible for study design and statistical analysis), and *domain experts* (who often provide the original research question and are responsible for data collection).
In smaller projects, analysts and domain experts might be the same persons. In these research settings, a broad review of the available literature typically plays a more important role.
Because most domain knowledge can only be expressed in unstandardized measurement units of a specific application, we argue that unstandardized effect sizes are usually preferable over standardized effect sizes for tailored simulation-based sample size planning. The basic idea of all strategies is to quantify or visualize the data-generating process implied by certain values of population parameters in an intuitive way that enables calibration of population parameters based on the available domain knowledge.
Although we use frequentist model estimation in our tutorial, many strategies demonstrated in this chapter are inspired by research on monitoring the plausibility of model assumptions [@gelmanBayesianWorkflow2020; @schadPrincipledBayesianWorkflow2021] and eliciting prior information [@bocktingSimulationbasedPriorKnowledge2024; @mikkolaPriorKnowledgeElicitation2023; @hartmannFlexiblePriorElicitation2020; @stefanPracticalChallengesMethodological2022] in applied Bayesian statistics.

@@

Tailored simulation-based sample size planning always requires access to domain knowledge. However, gathering domain knowledge and the specification of population parameter values is not always straightforward and can be better described as an iterative process, in which the plausibility of simulated data is repeatedly compared with the available domain expertise. As a first step, dedicated domain experts (if available) can be interviewed to *elicit* their domain knowledge about how the data of the planned study is expected to look. Because most domain experts are no experts in statistical modeling and GLMMs, they often struggle without further guidance to communicate their knowledge in a way that is useful when specifying the parameters for data simulation. For this reason, we suggest that after an initial interview, the analyst who is familiar with the structure of the GLMM selects an initial set of insightful descriptive statistics and model-based quantities. Then, they reenter into an iterative discussion with the domain experts where some set of population values are selected, and the plausibility of resulting implied quantities is evaluated. The population parameters are updated until the domain experts are satisfied with the result. During this process, the monitored descriptive statistics and model-based quantities can be updated or extended to capture as much available domain knowledge as possible.
Even if no dedicated domain experts are available, the basic principles stay the same except that initial parameter values are often derived from a more extensive literature review. Descriptive statistics and model-based quantities are still used to evaluate whether the implied assumptions about the data generating process are plausible. Often, this judgment evolves over several iterations as the analyst becomes more familiar with the model and the study design.^[Developing more effective strategies on how to elicit and visualize domain knowledge is currently an active area of research, primarily in the context of Bayesian modeling [@bocktingSimulationbasedPriorKnowledge2024; @mikkolaPriorKnowledgeElicitation2023; @stefanPracticalChallengesMethodological2022; @hartmannFlexiblePriorElicitation2020]. We expect that future advances in this field will also be highly relevant for tailored simulation-based sample size planning with frequentist models.]
